Large Matrix Manipulation for Unfolding Optimization

### 2019-04-30 NOvA/numu/unfolding_optimization/large_matrix_manipulation_v2.tex
Version two for the ND physics talk.

\begin{frame}{Overview}
  \begin{itemize}
    \item Currently, the number of iterations we use for unfolding was optimized in 1D neutrino energy only, but not in the full 3D $(\cos\theta_\mu,T_\mu,E_\nu)$ space.
    \item I have been trying to optimize the number of iterations for iterative unfolding in the full 3D space.
    \item In this study, RooUnfold is used to calculate the covariance matrix between bins.
    \item The total number of bins in $(\cos\theta_\mu,T_\mu,E_\nu)$ is $14\times 22\times 22=6776$. As a result, the covariance matrix is of size $6776\times 6776$.
    \item It turns out large matrices are hard to manipulate numerically. 
  \end{itemize}
  \bigskip
  \centering
  \textcolor{red}{Welcome to numerical linear algebra!}
\end{frame}

\begin{frame}{Visualizing a Covariance Matrix}
\begin{columns}
  \begin{column}{.6\textwidth}
    \includegraphics[width=\textwidth]{figures/cov_mat.eps}
  \end{column}
  \begin{column}{.4\textwidth}
    \begin{itemize}
      \item Log scale is used to make the structure more visible.
      \item Obviously this matrix is singular due to rows and columns with elements all zero. Those are unoccupied bins and unavoidable.
    \end{itemize}
  \end{column}
\end{columns}
\end{frame}

@@

@@

\begin{frame}{Condition Number and $\mathbf{V}_{jj}\mathbf{V}^{-1}_{jj}$}
  \begin{itemize}
    \item Consider the linear equation $\mathbf{A}\mathbf{x}=\mathbf{b}$. If small changes in $\mathbf{b}$ imply small changes in the solution $\mathbf{x}$, then $\mathbf{A}$ has a small condition number. Otherwise $\mathbf{A}$ has a big condition number.
    \item Not surprisingly, nearly singular matrices have large condition numbers.
    \item Since I don't know how to prove Eq.~(\ref{valid_v}) mathematically, I did what an experimentalist does: do some experiments.
    \item I generated random positive semidefinite matrices to simulate covariance matrices, and look at their condition numbers and $\mathbf{V}_{jj}\mathbf{V}^{+}_{jj}$'s.
  \end{itemize}
\end{frame}

@@

\begin{frame}{The Rank and Inverse of a Large Matrix Actually Depend on a Small Tolerance!}
  \begin{itemize}
    \item Intuitively, the number of valid $\rho_j$'s should not exceed the rank of the matrix. Therefore, knowing the rank of the matrix helps understand the problem.
    \item Turns out the rank of a large matrix depends on a small tolerance $\epsilon$. Singular values obtained by SVD (singular value decomposition) that are smaller than $\epsilon$ are set to exact zero to avoid numerical instability. This is called \textcolor{red}{low-rank approximation (or truncated SVD)}, and regarded as a way of regularizing the linear system.
    \item The same tolerance applies to obtaining the pseudoinverse of the matrix since the most common algorithm used to calculated the pseudoinverse is using SVD.
    \item I am going to show you the ranks, the $\rho_{avg}$'s among valid $\rho_j$'s of the covariance matrices as a function of number of iterations and compare results obtained with two different tolerance values.
  \end{itemize}
\end{frame}

@@

>>>>

\begin{frame}{Overview}
  \begin{itemize}
    \item Currently, the number of iterations we use for unfolding was optimized in 1D neutrino energy only, but not in the full 3D $(\cos\theta_\mu,T_\mu,E_\nu)$ space. We would like to re-optimize the number of iterations in the full 3D space.
    \item So far, I have been following other analyses to use RooUnfold for calculating the bin-to-bin covariance matrix.
    \item The total number of bins in $(\cos\theta_\mu,T_\mu,E_\nu)$ is $14\times 22\times 22=6776$. As a result, the covariance matrix is of size $6776\times 6776$.
    \item It turns out large matrices are hard to manipulate numerically.
    \item This study is done with \texttt{RyROOT}, \texttt{root\_numpy}, and \texttt{numpy.linalg}.
  \end{itemize}
  \bigskip
  \centering
  \textcolor{red}{Welcome to numerical linear algebra!}
\end{frame}

\begingroup
\footnotesize
\begin{frame}{Difficulties}
  \begin{enumerate}
    \item Using RooUnfold to calculate the covariance matrix takes exceedingly long time.
    \begin{itemize}
      \scriptsize
      \item \texttt{RooUnfold::Ereco()}\footnote{The only other public method I know of, \texttt{RooUnfold::GetMeasuredCov()}, gives me an empty matrix.} with option \texttt{RooUnfold::ErrorTreatment::kCovariance} is used to obtain the covariance matrix.
      \item We break down the time spent, and found doing more iterations takes no time, but retrieving the covariance matrix takes a full day. We think $10^8$ operations should be easy for modern computers.
      \item We might want to look at tens of different iterations, and the time spent will become unmanageable.
    \end{itemize}
    \item The metric for optimization involves a matrix inversion, and we found the covariance matrix are all nearly singular.
    \begin{itemize}
      \scriptsize
      \item When I calculate the inverse matrix, singular matrix warning messages are printed on the terminal reminding me of large numerical errors.
    \end{itemize}
  \end{enumerate}
  Strategy:
  \begin{enumerate}
    \item RooUnfold might be doing something inefficient. We plan to implement our own covariance matrix calculation.
    \item I looked into the singular matrix problem and summarize what I learned in this talk.
  \end{enumerate}
\end{frame}
\endgroup

\begin{frame}{Visualizing a Covariance Matrix}
\begin{columns}
  \begin{column}{.6\textwidth}
    \includegraphics[width=\textwidth]{figures/cov_mat.eps}
  \end{column}
  \begin{column}{.4\textwidth}
    \begin{itemize}
      \item Log scale is used to make the structure more visible.
      \item Obviously this matrix is singular due to rows and columns with elements all zero. Those are unoccupied bins and unavoidable with our binning.
    \end{itemize}
  \end{column}
\end{columns}
\end{frame}

@@

\begin{frame}{Geometric Meaning of Pseudoinverse}
  \begin{columns}
    \begin{column}{.5\textwidth}
      \includegraphics[width=.85\textwidth]{figures/pseudo_geo.png}
    \end{column}
    \begin{column}{.5\textwidth}
      \begin{itemize}
        \item Given a linear equation $\mathbf{Ax}=\mathbf{b}$, $\mathbf{A}^+\mathbf{b}$ is the least square solution to the problem. That is, $\mathbf{x}=\mathbf{A}^+\mathbf{b}$ minimizes $||\mathbf{Ax}-\mathbf{b}||^2$.
        \item Geometrically, it first project orthogonally the vector $\mathbf{b}$ to the subspace, range of $\mathbf{A}$, and invert from there.
        \item Algorithmically, it is obtained effortlessly from singular value decomposition (SVD).
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}{Condition Number\footnote{\footnotesize Smallness of $det(\mathbf{A})$ is not a good measure for invertibility. For example, $10^{-50}\mathbf{I_5}$ has extremely small determinant, but is readily invertible with a condition number $1$.} and $\mathbf{V}_{jj}\mathbf{V}^{-1}_{jj}$}
  \begin{itemize}
    \item Consider the linear equation $\mathbf{A}\mathbf{x}=\mathbf{b}$. If small changes in $\mathbf{b}$ imply small changes in the solution $\mathbf{x}$, then $\mathbf{A}$ has a small condition number. Otherwise $\mathbf{A}$ has a big condition number.
    \item Not surprisingly, nearly singular matrices have large condition numbers.
    \item Since I don't know how to prove Eq.~(\ref{valid_v}) mathematically, I did what an experimentalist does: do some experiments.
    \item I generated random positive semidefinite matrices to simulate covariance matrices, and look at their condition numbers and $\mathbf{V}_{jj}\mathbf{V}^{+}_{jj}$'s.
  \end{itemize}
\end{frame}

@@

\begin{frame}{The Rank and Inverse of a Large Matrix Actually Depend on a Small Tolerance!}
  \begin{itemize}
    \item Intuitively, the number of valid $\rho_j$'s should not exceed the rank of the matrix. Therefore, knowing the rank of the matrix helps understand the problem.
    \item Turns out the rank of a large matrix depends on a small tolerance $\epsilon$. Singular values obtained by SVD that are smaller than $\epsilon$ are set to exact zero to avoid numerical instability. This is called \textcolor{red}{low-rank approximation (or truncated SVD)}, and regarded as a way of regularizing the linear system.
    \item The same tolerance applies to obtaining the pseudoinverse of the matrix since the most common algorithm used to calculated the pseudoinverse is using SVD.
    \item I am going to show you the ranks, the $\rho_{avg}$'s among valid $\rho_j$'s of the covariance matrices as a function of number of iterations and compare results obtained with two different tolerance values.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Backup}

\end{frame}

\begin{frame}{My Working Knowledge on Singular Value Decomposition}
  \begin{columns}
    \begin{column}{.5\textwidth}
      \includegraphics[width=\textwidth]{figures/svd.png}
    \end{column}
    \begin{column}{.5\textwidth}
      \begin{itemize}
        \item $\mathbf{U}$ and $\mathbf{V}$ are unitary (orthogonal) matrices. $\boldsymbol{\Sigma}$ is diagonal with elements $\sigma_1\ge ... \ge\sigma_{min(m,n)}\ge 0$. $\sigma_i$ are called singular value.
        \item Rank of $\mathbf{A}$ equals the number of non-zero singular values.
        \item Pseudoinverse of $\mathbf{A}$ is $\mathbf{V}\boldsymbol{\Sigma}^+\mathbf{U}^T$. $\boldsymbol{\Sigma}^+$ is the reciprocal of each non-zero element on the diagonal of $\boldsymbol{\Sigma}$, leaving the zeros in place, and then transposing the matrix.
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}
